Microlensing as a possible probe of event-horizon structure in quasars

### 2016-07-20 paper1/ms.tex
editing intro

Modestly active galactic nucleii are present in the Galactic centre
and in M87.  In these two cases, the central engines ($<0.1\rm\,mas$
on the sky) are close to being resolved through very long baseline
interferometry, which shows preliminary indications of the
jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2013MNRAS.434..765K,2016arXiv160205527F}.
Current data do not deliver images but require fitting to predefined
models for the images.  A whole range of models have been applied,
starting from simple geometric models to more complex physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the data, and argued that the crescent is nothing but the
silhouette of the event horizon.

The great majority of quasars, however, lie at redshifts beyond 2
\citep{2014A&A...563A..54P} and their central engines would be orders
of magnitude smaller on the sky. The direct observations of the black
hole silhouettes of quasars are far beyond foreseeable instrumentation.
In the present paper, we consider a possible indirect method, related
to \cite{1999ApJ...524...49A}, through which gravitational
microlensing could probe the black hole shadow and its proximal quasar
environment.

Gravitational microlensing of quasars reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by an intervening
galaxies, without any changes in the intrinsic luminosity.
Microlensing affects only the light from the innermost part of the
quasar, such as the optical continuum
\citep[e.g.,][]{2012A&A...544A..62S} and is a consequence of two
things: the very small size of the central engine, and granularity of
the mass distribution of a lensing galaxy due to stars.  The latter
means that the local magnification is not a smooth function of source
position. It contains a complicated network of singular curves, known
as caustics.  Figure~\ref{fig:magnification_map} shows part of a
magnification map with a few caustics.  The lensed brightness would be
given by placing the source on such a magnification map and
integrating the surface brightness weighted by the magnification.
Most astrophysical sources straddle several caustics, and hence, their
net brightness varies smoothly with location.  The central engine of
quasars, however, is smaller than the typical spacing between
caustics.  As a result, the lensed brightness undergoes sudden changes
as a quasar crosses a caustic.  The effect supplies an upper limit on
the size of the central engine, and can also be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Caustics have an additional
remarkable feature: though they can be very complicated, they have
some universal properties well-known from catastrophe theory.  In
particular, very close to the simplest caustics (known as folds), the
magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/S1D_var_a.eps}
    \includegraphics[width = 0.48\textwidth]{figures/4avar_magnification.eps}
\caption{\label{fig:lightcurve_disk} Lightcurves of crescent sources with $R_n$ = 0.4 $R_p$.}
\end{figure}

@@

Figure 4 reveals that the lightcurve of a crescent source has more visible features than the other two light-curves
corresponding to the disc and gaussian shape. There are three regions where $S_{1D}$ and by inheritance, the lightcurve
 has distinguishable behaviour. The first region that would be recorded on a lightcurve plot represents the period
of time when the bright disk begins to be overlapped by the caustic and stops when just before the dark disk reaches
the caustic. During this period of time, the flux of light from the source is increasingly magnified. The second
period starts and ends with the overlapping of the dark disc. As a boundary of the two regions, there is a
distinguishable point where the slope of the magnification is drastically changed. This apparent discontinuity in the
first derivative of the magnification function is caused by the caustic amplification of the sudden drop in the $S_{1d}$
function. During the respective period, the magnification growth slows or even reverses during the first part of the
period and starts to grow faster again as the dark disk ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is infinite, which appears as a saddle point
on the lightcurve. Next, the final period corresponds to the case when the dark disc has cleared the caustic
and the bright disc continues to overlap with the caustic. During this period, the lightcurve reaches a peak
that for most of the parameter space is global and for the rest of the parameter space local.
At the end of the period, the growth of the magnification is negative infinite. the respective point appears 
as a second saddle point on the lightcurve. Past this point the magnification of any finite source will 
decay in roughly the same manner. \\

The impact of parameter $a$ on the shape of the lightcurve can be observed in figure 5 (a-var). Sources where 
the center of the dark disc reaches the caustic before the center of the bright disc are characterized by a smoother
broad peak in contrast to the cases where the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the caustic corresponds to larger and
larger magnifications until it becomes a local and even global peak. The effect of the $R_n$  parameter on the
light curve is presented in figures 6 and 7. For the particular set of parameters where $R_p = R_n +a$ the third
period of time discussed previously does not exist. Particularizing further, if the value of the radius of the dark
disc is comparable to the value of the radius of the bright one then the position and shape of the maximum magnification
 are strikingly different. In case the crescent reaches the caustic with the bright region first, the peak magnification
 happens when the dark region reaches the caustic and it is characterized by a sharp variation in magnification growth,
In the opposite case, the peak appears before the end of the bright disc reaches the caustic and shape is smoother. \\

@@

@@

\begin{figure}
% \centering
  \includegraphics[width=0.5\textwidth]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} Mock lightcurve fitted with the
  correct source type and caustic.  The upper panel shows the mock
  data points, and running through them is a seven-parameter best fit:
  these being the crescent parameters $(R_p,R_n,a,b)$ and additionally
  the start and end times and the overall normalisation.  (Ideal
  lightcurves for the same crescent parameters appear above in
  Figures~\ref{fig:lightcurve_gauss}, \ref{fig:lightcurve_disk},
  \ref{fig:lightcurve_crescent} and \ref{fig:char_points}.)  The curve
  has $\chi^2=243, P=0.32$ indicating a good fit.  The lower panel
  shows the posterior distributions for $r_{1/2}$ (solid histogram)
  and $R_n/R_p$ (dashed histogram).}
\end{figure}

@@

Nevertheless, the $b$ parameter is relevant for the calculation of $r_{1/2}$. 
It's qualitative effect is to decrease the value of the half-light radius when the absolute value of the parameter is increased.  
With two parameters constrained and another irrelevant to the shape of the lightcurve, two free parameters remain $R_n$ and $a$. 
Figure 4 reveals that the lightcurve of a crescent source has more visible features than the other two light-curves 
corresponding to the disc and Gaussian shape. The parameters $R_n$ and $a$ have strong influences on the shape of 
the microlensed lightcurve as can be seen in figures 5, 6 and 7. Moreover, the one-dimensional source profile corresponding to the direction perpendicular to the caustic reveal four characteristic points. The overlap of each of these points with the caustic leaves visible features on the lightcurve at the corresponding instances of time. In timely order, the instances correspond to the start of the overlap between the caustic and the bright disk, the start of the overlap between the caustic and the dark disk, the end of the overlap between the caustic and dark disk and finally the end of the overlap between the caustic and the bright disk.

>>>>

Modestly active galactic nucleii are present in the Galactic centre
and in M87.  In these two cases, the central engines ($<0.1\rm\,mas$
on the sky) are close to being resolved through very long baseline
interferometry, which shows preliminary indications of the
jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2016ApJ...820...90F}.
Current data do not deliver images but require fitting to predefined
models for the images.  A whole range of models have been applied,
starting from simple geometric models to more complex physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the available sub-mm interferometric data, They argued that the
crescent is nothing but the silhouette of the event horizon.

The great majority of quasars, however, lie at redshifts beyond 2
\citep{2014A&A...563A..54P} and their central engines would be orders
of magnitude smaller on the sky. The direct observations of the black
hole silhouettes of quasars are far beyond foreseeable
instrumentation.  In the present paper, we consider a possible
indirect method, related to \cite{1999ApJ...524...49A} and
\cite{2013ApJ...769..131C}, through which gravitational microlensing
could probe the black hole shadow and its proximal quasar environment.

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by an intervening
galaxies, without any changes in the intrinsic luminosity.  It is a
consequence of two things: the very small size of the central engine,
and granularity of the mass distribution of a lensing galaxy due to
stars.  The latter means that the local magnification is not a smooth
function of source position. It contains a complicated network of
singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence, their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  Quasar microlensing is mostly observed in the optical
continuum \citep[e.g.,][]{2012A&A...544A..62S} but is also known in
x-rays \citep[e.g.,][]{2016arXiv160207601N} and has been argued for at
cm wavelengths \citep{2000A&A...358..793K}.  The effect supplies an
upper limit on the size of the central engine
\citep[e.g.,][]{2015ApJ...814L..26M} and can be used to study the mass
distribution and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Microlensing caustics have an
additional remarkable feature: though they can be very complicated,
they have some universal properties well-known from catastrophe
theory.  In particular, very close to the simplest caustics (known as
folds), the magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/S1D_var_a.eps}
    \includegraphics[width = 0.48\textwidth]{figures/4avar_magnification.eps}
\caption{\label{fig:a_var} Lightcurves of crescent sources with $R_n$
  = 0.4 $R_p$.}
\end{figure}

@@

Figure 4 reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and gaussian shape. There are three regions where $S_{1D}$ and by
inheritance, the lightcurve has distinguishable behaviour. The first
region that would be recorded on a lightcurve plot represents the
period of time when the bright disk begins to be overlapped by the
caustic and stops when just before the dark disk reaches the
caustic. During this period of time, the flux of light from the source
is increasingly magnified. The second period starts and ends with the
overlapping of the dark disc. As a boundary of the two regions, there
is a distinguishable point where the slope of the magnification is
drastically changed. This apparent discontinuity in the first
derivative of the magnification function is caused by the caustic
amplification of the sudden drop in the $S_{1d}$ function. During the
respective period, the magnification growth slows or even reverses
during the first part of the period and starts to grow faster again as
the dark disk ends its overlap with the caustic. At the point where
the dark disc clears the caustic, the growth of the magnification is
infinite, which appears as a saddle point on the lightcurve. Next, the
final period corresponds to the case when the dark disc has cleared
the caustic and the bright disc continues to overlap with the
caustic. During this period, the lightcurve reaches a peak that for
most of the parameter space is global and for the rest of the
parameter space local.  At the end of the period, the growth of the
magnification is negative infinite. the respective point appears as a
second saddle point on the lightcurve. Past this point the
magnification of any finite source will decay in roughly the same
manner.

The impact of parameter $a$ on the shape of the lightcurve can be
observed in Figure~\ref{fig:a_var}. Sources where the center of the
dark disc reaches the caustic before the center of the bright disc are
characterized by a smoother broad peak in contrast to the cases where
the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the
caustic corresponds to larger and larger magnifications until it
becomes a local and even global peak. The effect of the $R_n$
parameter on the light curve is presented in
Figures~\ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}. For the particular set of
parameters where $R_p = R_n +a$ the third period of time discussed
previously does not exist. Particularizing further, if the value of
the radius of the dark disc is comparable to the value of the radius
of the bright one then the position and shape of the maximum
magnification are strikingly different. In case the crescent reaches
the caustic with the bright region first, the peak magnification
happens when the dark region reaches the caustic and it is
characterized by a sharp variation in magnification growth, In the
opposite case, the peak appears before the end of the bright disc
reaches the caustic and shape is smoother.

@@

\begin{figure}
\centering
  \includegraphics[width=0.5\textwidth]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} Mock lightcurve fitted with the
  correct source type and caustic.  The upper panel shows the mock
  data points, and running through them is a seven-parameter best fit:
  these being the crescent parameters $(R_p,R_n,a,b)$ and additionally
  the start and end times and the overall normalisation.  (Ideal
  lightcurves for the same crescent parameters appear above in
  Figures~\ref{fig:lightcurve_gauss}, \ref{fig:a_var},
  \ref{fig:lightcurve_crescent} and \ref{fig:char_points}.)  The curve
  has $\chi^2=243, P=0.32$ indicating a good fit.  The lower panel
  shows the posterior distributions for $r_{1/2}$ (solid histogram)
  and $R_n/R_p$ (dashed histogram).}
\end{figure}

@@

@@

Nevertheless, the $b$ parameter is relevant for the calculation of
$r_{1/2}$.  It's qualitative effect is to decrease the value of the
half-light radius when the absolute value of the parameter is
increased.  With two parameters constrained and another irrelevant to
the shape of the lightcurve, two free parameters remain $R_n$ and $a$.
Figure 4 reveals that the lightcurve of a crescent source has more
visible features than the other two light-curves corresponding to the
disc and Gaussian shape. The parameters $R_n$ and $a$ have strong
influences on the shape of the microlensed lightcurve as can be seen
in Figures~\ref{fig:a_var}, \ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}.  Moreover, the one-dimensional
source profile corresponding to the direction perpendicular to the
caustic reveal four characteristic points. The overlap of each of
these points with the caustic leaves visible features on the
lightcurve at the corresponding instances of time. In timely order,
the instances correspond to the start of the overlap between the
caustic and the bright disk, the start of the overlap between the
caustic and the dark disk, the end of the overlap between the caustic
and dark disk and finally the end of the overlap between the caustic
and the bright disk.
